UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-05 ms/main.tex
Revert "fix some images, add more stuff on FEV1, add citation on AFR/EUR FEV1 study"
This reverts commit 9f088e6fd106a716a22a0fbcb61aa0dfc3226ec4.

As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}). The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby. Using data about country of birth, we performed a similar analysis of a world map~\ref{fig:umap_world_map}, revealing subtle regional variation around the world.

@@

\subsubsection*{Patterns in phenotype related to population structure} 
Covariates such as height and leukocyte count (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure. Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and others have noticeably lower-than-average heights. More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals; an unpaired two sample t-test of self-identified White Irish and White British individuals reveals statistically significant differences in age-adjusted mean height between the populations, with British males being taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots). Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).

\subsubsection*{Comparing t-SNE and UMAP}

Identifying the best dimension reduction technique is to an extent subjective. However, we did find that using UMAP to preprocess the UKBB data allowed a standard t-SNE implementation to find representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

\subsection*{Discussion}
UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure. Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions. It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure. By patching together many local genetic neighbourhoods, the visualizations preserve aspects of both local and global structure, unlike any existing method that we are familiar with.

Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1KGP that were not visible through global PCA or Admixture analysis alone. With the UKBB data we revealed a range of patterns of covariation between geography, phenotypes and genotypes. Traits such as height showed continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showed sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.

We found that pre-processing the data with PCA allowed for considerable time savings. Groupings on ethnicity formed slowly as PCs were added until reaching a stable number around 10 to 15 PCs. Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}. In our experience, we find it useful to include as many PCs as are available and computationally feasible.

Like most non-linear methods, UMAP lacks direct interpretability. It is premised on local distances being more important than global distances; while points that appear very close locally can be thought to be more similar to each other, points that are distant from each other are not necessarily very different from each other. Figure~\ref{fig:supp_megamontage_pc10_50} demonstrates that disconnected population clusters can appear at different locations depending on parameter choices. For this reason, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.

UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and exaggerates the genetic variation within a population, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

>>>>

As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}). The geographical patterns outside major urban centers are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby. Using data about country of birth, we performed a similar analysis of a world map~\ref{fig:umap_world_map}, revealing subtle regional variation around the world.

@@

\subsubsection*{Identifying patterns in phenotype variation related to genetic population structure}

Covariates such as height and leukocyte count (figure~\ref{fig:umap_height_female}) and autoimmune and asthma-related measures (figures~\ref{fig:supp_ukbb_basophill_f} to \ref{fig:supp_ukbb_neutrophill_m}) correlate strongly with both discrete and continuous population structure. Several populations in figure~\ref{fig:umap_height_female}, including South Asian, East Asian, African, and others have noticeably lower-than-average heights. More subtle patterns are also visible: the area of the projection in figures~\ref{fig:umap_ukbb_pc10} with the cluster of White Irish people appears more blue than the main body of White British individuals. To quantify and statistically test these qualitative observations, we performed an unpaired two sample t-test of self-identified White Irish and White British individuals and found British males taller on average by 0.846cm (p-value $2.10\times 10^{-23}$) and British females by 0.763cm (p-value $3.65\times 10^{-23}$) (see figures~\ref{fig:supp_box_height_f} and \ref{fig:supp_box_height_m} for boxplots). Height differences between Irish and British populations have been previously observed but the direction of the difference is not consistent\cite{robinson2015population,komlos1994stature}.

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations \sgcomment{point to box plots again?}.

Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan \sgcomment{if test is not about those born in Japan, needs to be clarified!} have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males (see figure~\ref{fig:supp_fev_ridgeplots} for distributions). For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.

Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups. This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.

\subsubsection*{Comparing t-SNE and UMAP}
\sgcomment{tweaked this a bit -- would be good to include in reply to reviewer}
Identifying the best dimension reduction technique is challenging, both because the `best' representation depends on context, and because convergence issues may mean that a good theoretical model for dimensional reduction might perform poorly because of challenges in numerical optimization. To assess whether the relatively poor performance of t-SNE could be due to convergence rather than a flawed model, we used UMAP to preprocess the UKBB data and provide a starting point to a standard t-SNE implementation. This led to representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

\subsection*{Discussion}
UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure. It is easy to use and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions. It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure. By patching together many local genetic neighbourhoods, the visualizations preserve aspects of both local and global structure, unlike any existing method that we are familiar with.

Using UMAP reveals  clusters that would have been difficult to identify via pairwise PCA plots or Admixture analysis, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1KGP. More importantly, UMAP helps reveal patterns of covariation between geography, phenotypes and genotypes. Traits such as height showed continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showed sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.

We found that pre-processing the data with PCA allowed for time savings, but identifying an optimal number of PCs to use is challenging. Groupings on ethnicity formed slowly as PCs were added until reaching a stable number around 10 to 15 PCs. Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.

\subsubsection*{Caveats}
The existence of a (small) number of adjustable parameter to UMAP, including PC cutoff and number of neighbors, opens the door to tweaking. Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices. In the absence of clear theoretical rationales, we generally recommend to use as many PCs as are available and computationally feasible, and to use the default number of neighbours (15). We sometimes found that lower number of PCs led to a simpler shape that facilitated discussion (e.g., figure \ref{fig:fig_ukbb}). We recommend reporting on a range of parameter values and following up on observations with statistical testing. Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations.

UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots has a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.
